Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-15 submission/manuscript.Rmd
shigeki edits results

We found a large proportion of genes that are significantly upregulated within the _Kdm5c_-KO brain (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) are enriched in non-brain tissues in wild-type mice (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B). The majority of tissue-enriched differentially expressed genes (tissue-enriched DEGs) were testis genes (Figure 1A-C). Even though the testis has the largest total number of tissue-biased genes compared to any other tissue (`r gettissval("count", "Testis")` genes), testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Testis"), 3)`, Fisher's Exact Test). One example of a testis-enriched gene misexpressed in the _Kdm5c_-KO brain is _FK506 binding protein 6 (Fkbp6)_, a known regulator of piRNA expression and meiosis in germ cells[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012] (Figure 1C).

@@

To further validate if these testis DEGs are truly germline genes, we then assessed their expression in somatic versus germ cells within the testis. We first compared their expression in wild-type testes to those without germ cells[@muellerIndependentSpecializationHuman2013], which was accomplished by heterozygous _W_ and _Wv_ mutations in the enzymatic domain of _c-Kit_ (Kit^W/Wv^)[@handelSertoliCellDifferentiation1979]. Almost all _Kdm5c_-KO testis-enriched DEGs lost expression with germ cell depletion (Figure 2B). We then assessed testis-enriched DEG expression in a published single cell RNA-seq dataset that identified cell type-specific markers within the testis[@greenComprehensiveRoadmapMurine2018]. Some _Kdm5c_-KO testis-enriched DEGs were classified as specific markers for different germ cell types (e.g. spermatogonia, spermatocytes, round spermatids, and elongating spermatids), yet none marked somatic cells (Figure 2C). Together, these data demonstrate that the _Kdm5c_-KO brain aberrantly expresses germline genes, but not somatic testicular genes, reflecting an erosion between somatic versus germline identity.

@@

We first identified _Kdm5c_-KO EpiLC DEGs in our previously published RNA-seq dataset[@samantaActivationXistEvolutionarily2022] (DESeq2, log2 fold change > `r l2fcco`, q < 0.1) and then assessed tissue-enriched gene expression. Similar to the _Kdm5c_-KO brain, we observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3B). Markers of primed pluripotency, such as _Dnmt3b_, _Fgf5_, _Pou3f1_, and _Otx2_, were unaltered in _Kdm5c_-KO EpiLCs (Figure 3C). Furthermore, _Kdm5c_-KO cell morpholgy during ESC to EpiLC differentiation (Figure 3D) appeared normal, indicating KDM5C loss does not impair EpiLC formation.

To determine if germline DEGs are constitutively dysregulated or if they can change over the course of development, we next compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain. We found the majority of germline DEGs were unique to either EpiLCs or the brain, with only _Cyct_ shared across all tissue/cell types (Figure 3E-F). EpiLCs had particularly high enrichment of meiosis-related gene ontologies (Figure 3G), such as meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO:0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`). While there was modest enrichment of meiotic gene ontologies in both brain regions, the _Kdm5c_-KO hippocampus primarily expressed late-stage spermatogenesis genes, such those involved in sperm axoneme assembly (GO:0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`) and sperm motility (GO:0097722, p.adjust = `r signif(getGO("sperm motility", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`).

Notably, DEGs unique to _Kdm5c_-KO EpiLCs included key drivers of germline identity, such as _Stimulated by retinoic acid 8 (Stra8)_ and _Deleted in azoospermia like (Dazl)_ (Figure 3H). These genes are typically expressed during embryonic germ cell development to commit PGCs to the germline fate, but are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019]. Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo[@macfarlanEmbryonicStemCell2012; @guptaGenomewideScreenReveals2023]. However, we did not see misexpression of two-cell embryo-specific genes, like _Duxf3 (Dux)_ (q = `r signif(duxf3, 3)`) and _Zscan4d_ (q = `r signif(zscan4d, 3)`), indicating _Kdm5c_-KO in EpiLCs do not revert back to a 2-cell-like state (Figure 3H). Altogether, these results demonstrate that while the _Kdm5c_-KO brain primarily expresses spermiogenesis genes, _Kdm5c_-KO EpiLCs express key drivers of germline identity and meiosis.

@@

It is currently unknown if the misexpression of germline genes is influenced by chromosomal sex, as previous studies on germline gene repressors have focused on their regulation in males[@pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021]. We explored the impact of sex upon germline gene misexpession by comparing their dysregulation in male _Kdm5c_ hemizygous knockout (XY _Kdm5c_-KO), female homozygous knockout (XX _Kdm5c_-KO) and female heterozygous (XX _Kdm5c_-HET) EpiLCs. [@samantaActivationXistEvolutionarily2022]. We first identified differentially expressed genes (DEGs) compared to sex-matched wild-type controls (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco_ESCEpiLC`, q < 0.1) and then filtered for germline-enriched genes.

Homozygous and heterozygous females expressed over double the number of germline-enriched genes than hemizygous males (Figure 4A). While the majority of germline DEGs in _Kdm5c_-KO males were also dysregulated in females (74%), there were also many male-specific and female-specific germline DEGs (Figure 4B). We the compared the known functions of germline genes dysregulated only in females (XX only - XX _Kdm5c_-KO, XX _Kdm5c_-HET, or both), only in males (XY only), or in all samples (shared). Female-specific germline DEGs were enriched for meiotic (GO:XXX meiotic cell cycle) and flagellar (GO:XXX cilium movement) functions, male-specifc DEGs had mitochondrial and cell signaling gene functions (GO:XXX protein localization to mitochondrion). Germline transcripts expressed in both sexes were enriched for meiotic (meiotic nuclear division) and egg-specific functions (female gamete generation).

The majority of germline genes expressed in both sexes had a greater log2 fold change from wild-type in females compared to males (Figure 4D). This increased degree of dysregulation in females, along with the increased total number of germline genes, indicates females are more sensitive to the loss of KDM5C-mediated suppression of germline genes. Female sensitivity could be due to improper X chromosome inactivation (XCI) in _Kdm5c_ mutants, as many spermatogenesis genes lie on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004]. However, both shared and female-specific germline DEGs were not biased towards the X chromosome, with the majority of genes instead lying on autosomes (Figure 4G). Thus, while female EpiLCs are more prone to germline gene misexpression with KDM5C loss, it is likely independent of potential defects in XCI.

@@

While many germline genes act in both the male and female germline, some display sex-biased expression or have functions unique to eggs and sperm. To comprehensively assess whether germline gene sex corresponds with _Kdm5c_ mutant sex, we filtered our list of germline-enriched genes for egg (XX) and sperm (XY) biased genes. We defined sex-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex. This yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes (Figure 4H), which is consistent with the testis overall having a more unique transcriptome than the ovary[@liComprehensiveMouseTranscriptomic2017]. We found egg, sperm, and unbiased germline genes were dyregulated in all _Kdm5c_ mutants (Figure 4I-J). Germline genes dysregulated exclusively in males and females were also not biased towards their corresponding germ cell sex. This indicates differences between male and female germline gene dysregulation is not due to sex-specific activation of sperm or egg-specific transcriptional programs. These results demonstrate that sex influences the degree of germline gene misexpression with KDM5C loss, but not the germ cell sex from which the germline genes originate.

@@

Previous work suggests KDM5C represses germline genes during early development, as re-expression of KDM5C in knockout neuronal cultures fails to suppress two hippocampal germline DEGs[@scandagliaLossKdm5cCauses2017]. KDM5C binds to several germline gene promoters in embryonic stem cells (ESCs) but is lost in neurons[@scandagliaLossKdm5cCauses2017]. However, the lack of a comprehensive list of germline-enriched genes prohibited systematic characterization of KDM5C binding at germline gene promoters. It is currently unclear if KDM5C is enriched at germline gene promoters, what types of germline genes KDM5C regulates, and if its binding is maintained at any germline genes in neurons.

@@

The majority of promoters bound by KDM5C in PNCs were also bound in EpiLCs (513 shared promoters), however a large portion of gene promoters were only bound by KDM5C in EpiLCs (3677 EpiLC only promoters) (Figure 5B). Genes bound by KDM5C in both PNCs and EpiLCs were enriched for functions involving nucleic acid turnover, such as deoxyribonucleotide metabolic process (GO:0009262, p.adjust = `r signif(getGO("deoxyribonucleotide metabolic process", GO_KDM5C_ChIP, 'p.adjust', "Shared"), 3)`) (Figure 5C). Germline-specific ontologies were only enriched in promoters unique to EpiLCs, such as meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) and meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) (Figure 5C). There were no ontologies significantly enriched for genes only bound by KDM5C in PNCs. We next evaluated KDM5C binding around the transcription start site (TSS) of all germline-enriched genes. In EpiLCs, We observed modest KDM5C signal at about half of all germline genes (Figure 5D). Based on our ChIP-seq peak cut-off criteria, KDM5C was strongly bound to about 10% of germline gene promoters in EpiLCs (Figure 5E). In condordance with our gene ontology results, we did not observe KDM5C accumiulation at any germline gene promtoers in PNCs (Figure 5D). Together, these results demonstrate KDM5C is significantly enriched at a subset of germline gene promoters in EpiLCs, including meiotic genes, but does not directly regulate germline genes in neurons.

Many germline-specific genes are suppressed by the transcription factor heterodimers E2F6/DP1 and MGA/MAX, which respectively bind E2F and E-box motifs[@hurlinMgaDualspecificityTranscription1999; @pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @stielowMGAL3MBTL2E2F62018; @dahletE2F6InitiatesStable2021]. To elucidate if KDM5C is recruited to germline gene promoters by a similar mechanism, we used HOMER to identify transcription factor motifs enriched at KDM5C-bound or unbound germline gene promoters[@heinzSimpleCombinationsLineageDetermining2010] (TSS +/- 500 bp, q-value < 0.1). MAX and E2F6 binding sites were significantly enriched at germline genes bound by KDM5C in EpiLCs, but not at germline genes unbound by KDM5C (MAX q-value: 0.0068, E2F6 q-value: 0.0673, E2F q-value: 0.0917) (Figure 5F). One third of KDM5C-bound promoters contained the consensus sequence for either E2F6 (E2F, 5'-TCCCGC-3'), MGA (E-box, 5'-CACGTG-3'), or both, but only 17% of KDM5C-unbound genes contained these motifs (Figure 5G). KDM5C-unbound germline genes were intstead enriched for multiple RFX transcription factor binding sites (RFX q-value < 0.0001, Rfx2 q-value < 0.0001, Rfx5 q-value < 0.0001) (Figure 5H). RFX transcription factors bind X-box motifs[@gajiwalaStructureWingedhelixProtein2000] to promote ciliogenesis[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009]. Enriched RFX trancription factors included RFX2, a central regulator of post-meiotic spermatogeneis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016]. Interestingly, RFX2 mRNA is upregualted in _Kdm5c_-KO EpiLCs (Figure 5I), but is not a direct target of KDM5C. Thus, distinct transcription factor programs regulate the expression of KDM5C-bound and unbound germline genes.

Finally, we compared KDM5C binding at the promoters of RNA-seq DEGs to determine if the germline genes transcribed in _Kdm5c_-KO cells are direct targets of KDM5C (Figure 5J-L). In EpiLCs, KDM5C was bound to about one third of EpiLC-specific and brain-specific germline DEG promoters (EpiLC only: `r 18/50*100`%, Brain only: `r signif(7/21*100,3)`%). Again, we did not observe any KDM5C binding at germline gene promoters in PNCs, even for brain-specific DEGs. Some notable differences in KDM5C binding for EpiLC-specific DEGs included _Dazl_ and _Stra8_, two key drivers of germline identity discussed above. Although both mRNAs are expressed in _Kdm5c_-KO EpiLCs, KDM5C is only bound to _Dazl_'s promoter and not _Stra8_'s in EpiLCs (Figure 5K). In contrast to the unique DEGs, 3 out of the 4 genes dysregulated in both the brain and EpiLCs are bound by KDM5C around the TSS. Altogether, this suggests KDM5C decommssions a subset of germline genes in EpiLCs but is no longer recruited to germline gene promoters in neurons. Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C recruitment to their promoters.

>>>>

We found a large proportion of genes that are significantly upregulated within the _Kdm5c_-KO brain (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) are enriched in non-brain tissues in wild-type mice (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B). The majority of tissue-enriched differentially expressed genes (tissue-enriched DEGs) were testis genes (Figure 1A-C). Even though the testis has the largest total number of biased genes (`r gettissval("count", "Testis")` genes) compared to any other tissue, testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Testis"), 3)`, Fisher's Exact Test). One example of a testis-enriched gene misexpressed in the _Kdm5c_-KO brain is _FK506 binding protein 6 (Fkbp6)_, a known regulator of piRNA expression and meiosis[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012] (Figure 1C).

@@

To further validate if these testis DEGs are truly germline genes, we assessed their expression in somatic versus germ cells within the testis. We first compared their expression in wild-type testes to those without germ cells[@muellerIndependentSpecializationHuman2013], which was accomplished by heterozygous _W_ and _Wv_ mutations in the enzymatic domain of _c-Kit_ (Kit^W/Wv^)[@handelSertoliCellDifferentiation1979]. Almost all _Kdm5c_-KO testis-enriched DEGs lost expression with germ cell depletion (Figure 2B). We then assessed testis-enriched DEG expression in a published single cell RNA-seq dataset that identified cell type-specific markers within the testis[@greenComprehensiveRoadmapMurine2018]. Some _Kdm5c_-KO testis-enriched DEGs were classified as specific markers for different germ cell types (e.g. spermatogonia, spermatocytes, round spermatids, and elongating spermatids), yet none marked somatic cells (Figure 2C). Together, these data demonstrate that the _Kdm5c_-KO brain aberrantly expresses germline genes, but not somatic testicular genes, reflecting an erosion of the soma-germline boundary.

@@

We first identified _Kdm5c_-KO EpiLC DEGs in our previously published RNA-seq dataset[@samantaActivationXistEvolutionarily2022] (DESeq2, log2 fold change > `r l2fcco`, q < 0.1) and then assessed tissue-enriched gene expression. Similar to the _Kdm5c_-KO brain, we observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3B). Markers of primed pluripotency, such as _Dnmt3b_, _Fgf5_, _Pou3f1_, and _Otx2_, were unaltered in _Kdm5c_-KO EpiLCs (Figure 3C). Furthermore, _Kdm5c_-KO cell morpholgy during ESC to EpiLC differentiation appeared normal (Figure 3D), indicating KDM5C loss does not impair EpiLC formation.

We next compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain to determine if germline DEGs are constitutively dysregulated or change over the course of development. We found the majority of germline DEGs were unique to either EpiLCs or the brain, with only _Cyct_ shared across all tissue/cell types (Figure 3E-F). EpiLCs had particularly high enrichment of meiosis-related gene ontologies (Figure 3G), such as meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO:0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`). While there was modest enrichment of meiotic gene ontologies in both brain regions, the _Kdm5c_-KO hippocampus primarily expressed late-stage spermatogenesis genes involved in sperm axoneme assembly (GO:0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`) and sperm motility (GO:0097722, p.adjust = `r signif(getGO("sperm motility", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`).

Notably, DEGs unique to _Kdm5c_-KO EpiLCs included key drivers of germline identity, such as _Stimulated by retinoic acid 8 (Stra8)_ and _Deleted in azoospermia like (Dazl)_ (Figure 3H). These genes are typically expressed during embryonic germ cell development to commit PGCs to the germline fate, but are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019]. Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo[@macfarlanEmbryonicStemCell2012; @guptaGenomewideScreenReveals2023]. However, we did not see misexpression of two-cell stage-specific genes, like _Duxf3 (Dux)_ (q = `r signif(duxf3, 3)`) and _Zscan4d_ (q = `r signif(zscan4d, 3)`), indicating _Kdm5c_-KO in EpiLCs do not revert back to a 2-cell-like state (Figure 3H). Altogether, these results demonstrate that while the _Kdm5c_-KO brain primarily expresses spermiogenesis genes, _Kdm5c_-KO EpiLCs express key drivers of germline identity and meiosis.

@@

It is currently unknown if the misexpression of germline genes is influenced by sex, as previous studies on germline gene repressors have focused on males[@pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021]. We explored the impact of sex upon germline gene misexpession by comparing their dysregulation in male _Kdm5c_ hemizygous knockout (XY _Kdm5c_-KO), female homozygous knockout (XX _Kdm5c_-KO), and female heterozygous knockout (XX _Kdm5c_-HET) EpiLCs. [@samantaActivationXistEvolutionarily2022].

Homozygous and heterozygous females expressed over double the number of germline-enriched genes than hemizygous males (Figure 4A). While the majority of germline DEGs in _Kdm5c_-KO males were also dysregulated in females (74%), there were also many male-specific and female-specific germline DEGs (Figure 4B), such as _Tktl2_ and _Esx1_. We the compared the known functions of germline genes dysregulated only in females (XX only - XX _Kdm5c_-KO, XX _Kdm5c_-HET, or both), only in males (XY only), or in all samples (shared). Female-specific germline DEGs were enriched for meiotic (GO:0051321 meiotic cell cycle) and flagellar (GO:0003341 cilium movement) functions, male-specifc DEGs had mitochondrial and cell signaling gene functions (GO:0070585 protein localization to mitochondrion). Germline transcripts expressed in both sexes were enriched for meiotic (GO:0140013 meiotic nuclear division) and egg-specific functions (GO:0007292 female gamete generation).

The majority of germline genes expressed in both sexes had a greater log2 fold change in females compared to males (Figure 4D). This increased degree of dysregulation in females, along with the increased total number of germline genes, indicates females are more sensitive to losing KDM5C-mediated suppression of germline genes. Female sensitivity could be due to improper X chromosome inactivation (XCI) in _Kdm5c_ mutants[@samantaActivationXistEvolutionarily2022], as many spermatogenesis genes lie on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004]. However, both shared and female-specific germline DEGs were not biased towards the X chromosome and the majority of female DEGs instead lie on autosomes (Figure 4G). Thus, while female EpiLCs are more prone to germline gene misexpression with KDM5C loss, it is likely independent of potential defects in XCI.

@@

While many germline genes act in both the male and female germline, some display sex-biased expression or have functions unique to eggs and sperm. Therefore, we wondered if _Kdm5c_ mutant males express sperm genes while mutant females express egg genes. To comprehensively assess whether germline gene sex corresponds with _Kdm5c_ mutant sex, we filtered our list of germline-enriched genes for egg and sperm-biased genes. We defined germ cell sex-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex. This yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes (Figure 4H). We found egg, sperm, and unbiased germline genes were dyregulated in all _Kdm5c_ mutants (Figure 4I-J). Germline genes dysregulated exclusively in_Kdm5c_ mutant males and females were also not biased towards their corresponding germ cell sex. This indicates sex differences in germline gene dysregulation is not due to sex-specific activation of sperm or egg transcriptional programs. These results demonstrate that the sex of _Kdm5c_ mutant cells influences the degree of germline gene, independent of germline gene sex.

- **note: the edited last sentence ended with "independent of germ cell sex", but I think if it's read out of context that sounds like we tested KDM5C in the four core genotypes. But I'm not sure if "germline gene sex" is confusing**

@@

KDM5C binds to the promoters of several germline genes in embryonic stem cells (ESCs) but its binding is absent in neurons[@scandagliaLossKdm5cCauses2017]. However, the lack of a comprehensive list of germline-enriched genes prohibited systematic characterization of KDM5C binding at germline gene promoters. Thus, it is unclear if KDM5C is enriched at germline gene promoters, what types of germline genes KDM5C regulates, and if its binding is maintained at any germline genes in neurons.

@@

The majority of promoters bound by KDM5C in PNCs were also bound in EpiLCs (513 shared promoters), however a large portion of gene promoters were only bound by KDM5C in EpiLCs (3677 EpiLC only promoters) (Figure 5B). Genes bound by KDM5C in both PNCs and EpiLCs were enriched for functions involving nucleic acid turnover, such as deoxyribonucleotide metabolic process (GO:0009262, p.adjust = `r signif(getGO("deoxyribonucleotide metabolic process", GO_KDM5C_ChIP, 'p.adjust', "Shared"), 3)`) (Figure 5C). Germline-specific ontologies were only enriched in promoters unique to EpiLCs, such as meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) and meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) (Figure 5C). There were no ontologies significantly enriched for genes only bound by KDM5C in PNCs. We next evaluated KDM5C binding around the transcription start site (TSS) of all germline-enriched genes. In EpiLCs, we observed modest KDM5C signal at about half of all germline genes (Figure 5D). Based on our ChIP-seq peak cut-off criteria, KDM5C was strongly bound to about 10% of germline gene promoters in EpiLCs (Figure 5E). One notable gene that lacked KDM5C binding was _Stra8_, even though its mRNA is expressed in _Kdm5c_-KO EpiLCs (Figure 5K). In EpiLCs, KDM5C was only bound to about one third of _Kdm5c_-KO RNA-seq DEG promoters (EpiLC only DEGs: `r 18/50*100`%, Brain only DEGs: `r signif(7/21*100,3)`%), but 3 out of the 4 genes dysregulated in both the brain and EpiLCs (Supplementary figure XXX). In condordance with our gene ontology results, we did not observe KDM5C accumulation at any germline gene promtoers in PNCs (Figure 5D). Together, these results demonstrate KDM5C is recruited to a subset of germline genes in EpiLCs, including meiotic genes, but does not directly regulate germline genes in neurons. Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C binding to their promoters.

Many germline-specific genes are suppressed by the transcription factor heterodimers E2F6/DP1 and MGA/MAX, which respectively bind E2F and E-box motifs[@hurlinMgaDualspecificityTranscription1999; @pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @stielowMGAL3MBTL2E2F62018; @dahletE2F6InitiatesStable2021]. To elucidate if KDM5C is recruited to germline gene promoters by a similar mechanism, we used HOMER to identify transcription factor motifs enriched at KDM5C-bound or unbound germline gene promoters[@heinzSimpleCombinationsLineageDetermining2010] (TSS +/- 500 bp, q-value < 0.1). MAX and E2F6 binding sites were significantly enriched at germline genes bound by KDM5C in EpiLCs, but not at germline genes unbound by KDM5C (MAX q-value: 0.0068, E2F6 q-value: 0.0673, E2F q-value: 0.0917) (Figure 5F). One third of KDM5C-bound promoters contained the consensus sequence for either E2F6 (E2F, 5'-TCCCGC-3'), MGA (E-box, 5'-CACGTG-3'), or both, but only 17% of KDM5C-unbound genes contained these motifs (Figure 5G). KDM5C-unbound germline genes were intstead enriched for multiple RFX transcription factor binding sites (RFX q-value < 0.0001, RFX2 q-value < 0.0001, RFX5 q-value < 0.0001) (Figure 5H). RFX transcription factors bind X-box motifs[@gajiwalaStructureWingedhelixProtein2000] to promote ciliogenesis[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009] and among them is RFX2, a central regulator of post-meiotic spermatogeneis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016]. Interestingly, RFX2 mRNA is derepressed in _Kdm5c_-KO EpiLCs (Figure 5I), however it is not a direct target of KDM5C (Supplementary figure XXX). Thus, RFX2 is a candidate transcription factor for driving the ectopic expression of KDM5C-unbound germline genes in _Kdm5c_-KO cells.
